---
title: "PreliminaryData"
author: "Melissa Lowe"
date: "1/29/2019"
output: pdf_document
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE, fig.path='prelimfigs/')
```
Import Data:

```{r}
library(haven)
mort_adjud <- read_dta("~/Desktop/COPDGeneData/Mortality_Jan2019.dta", 
    NULL)
```
```{r}
#copdgene_p1p2_all <- read_sas("~/Desktop/COPDGeneData/copdgene_p1p2_all_visit_30dec18.sas7bdat",NULL)
```
```{r}
copdgene_p1p2_flat <- read_dta("~/Desktop/COPDGeneData/P1P2_Pheno_Flat_All_sids_Dec18.dta", 
    NULL)
```


Create my own dataset of variables of interest:

We know we'll need sid, visit num, ccenter, visit date, gender, race, smoking status, age at baseline, visit type, exclude lungtrans, height, wegiht, distwalked, cigperdaysmoknow, all copdexac, lungproc_lungtransplant, copdafe, emphage, smokstartage, ats_packyears, yearssincequit, fev1pp_utah, fev1_fvc_utah, 

```{r}

#need to deal with the atomic labels which are actually storing the subject ids
library(sjlabelled)

#save the old sid values just in case
copdgene_p1p2_flat$sid2 <- copdgene_p1p2_flat$sid

#pull the labels as necessary
copdgene_p1p2_flat$sid <- get_labels(copdgene_p1p2_flat$sid)

#make them both into characters so they are the same. 
mort_adjud$sid <- as.character(mort_adjud$sid)


library(tidyverse)


#subset so I don't have an insanely large data set


flat_1 <- copdgene_p1p2_flat %>% select(sid, gender,race,Visit_Date_P1,ccenter_P1, EverSmokedCig_P1, smoking_status_P1, Age_P1, Height_CM_P1, Weight_KG_P1, BMI_P1, distwalked_P1, ATS_PackYears_P1,YearsSinceQuit_P1,Duration_Smoking_P1,Severe_Exacerbations_P1, Exacerbation_Frequency_P1, finalGold_P1, FEV1_FVC_utah_P1, FEV1_utah_P1, FVC_utah_P1, pctEmph_Thirona_P1, AWT_seg_Thirona_P1, pctGasTrap_Thirona_P1, Visit_Date_P2, ccenter_P2, years_from_baseline, smoking_status_P2, Age_P2, VisitType_P2, Pi10_Thirona_P1, Height_CM_P2, Weight_KG_P2,                
BMI_P2,distwalked_P2,SmokStopAge_P2,ATS_PackYears_P2, YearsSinceQuit_P2, Severe_Exacerbations_P2, Exacerbation_Frequency_P2, finalGold_P2, FEV1_FVC_utah_P2, FEV1_utah_P2, FVC_utah_P2, Days_Phase1_Phase2, Change_P1_P2_Smoking_Status, smoking_status_change, pctEmph_Thirona_P2, Pi10_Thirona_P2, pctGasTrap_Thirona_P2, AWT_seg_Thirona_P2)

#possibly useful data set
mort_1 <- mort_adjud %>% select(sid, vital_status, mortality_survival_vetted, mortality_survival_vital_status, months_followed_net, months_followed_visits, months_followed_LFU, months_followed_SSDI_search, days_followed, mortality_cohort)

#number of days since January 1, 1960 is the date.

flat_1$followdays <- flat_1$Visit_Date_P2 - flat_1$Visit_Date_P1

hist(flat_1$followdays)

mean(na.exclude(flat_1$followdays))
summary(na.exclude(flat_1$followdays))

flat_1$diff_spiro_ratio <- flat_1$FEV1_FVC_utah_P2 -  flat_1$FEV1_FVC_utah_P1

flat_1$diff_spiro_fev1 <- flat_1$FEV1_utah_P2 - flat_1$FEV1_utah_P1

flat_1$diff_emphys <- flat_1$pctEmph_Thirona_P2 - flat_1$pctEmph_Thirona_P1

flat_1$diff_gas_trap <- flat_1$pctGasTrap_Thirona_P2 - flat_1$pctGasTrap_Thirona_P1

flat_1$diff_Pi10 <-  flat_1$Pi10_Thirona_P1 - flat_1$Pi10_Thirona_P2

flat_1$diff_AWT <- flat_1$AWT_seg_Thirona_P1 - flat_1$AWT_seg_Thirona_P2


flag_items <- as.data.frame(cbind(flat_1$diff_spiro_ratio,flat_1$diff_spiro_fev1, flat_1$diff_emphys, flat_1$diff_gas_trap, flat_1$diff_Pi10, flat_1$diff_AWT))

names(flag_items) <- c("diff_ratio","diff_fev1", "diff_emph", "diff_gastrap", "diff_pi10", "diff_awt")


```

Summary of Visits:




```{r, include=FALSE}
summaries <- function(x) {
  y <- na.exclude(x)
  minimum = round(min(y), digits =2)
  #med = median(y)
  maximum = round(max (y), digits=2)
  mu = round(mean(y), digits =2 )
  stdev = round(sd(y), digits=2)
  quant <- round(quantile(y, c(.05, .5, .95)), digits =3)
  N =  length(y)
  allsummary <- c(mu, stdev, N, minimum, maximum, quant)
  names(allsummary) <- c("Mean", "SD", "N", "Minimum", "Maximum", ".05,.5, .95")
  return(allsummary)
}
```


Summary Tables of Outcomes of Interest
```{r}
outcome_table <- as.data.frame(t(sapply(flag_items, summaries)))
outcome_table

```

Currently, they're using the 95 percentile to mark where they think a serious change in disease status would be for these markers. 

This is obviously a fairly artificial marker. 
```{r}

flat_1$flag_spiro_ratio <- ifelse(flat_1$diff_spiro_ratio >= 0.09, 1, 0) 

sum(na.exclude(flat_1$flag_spiro_ratio))

flat_1$flag_fev1 <-  ifelse(flat_1$diff_spiro_fev1 >= 0.218, 1, 0)

sum(na.exclude(flat_1$flag_fev1))

flat_1$flag_emphys <-  ifelse(flat_1$diff_emphys >= 7.023, 1, 0)

sum(na.exclude(flat_1$flag_emphys))

flat_1$flag_gastrap <-  ifelse(flat_1$diff_gas_trap >= 17.165, 1, 0)

sum(na.exclude(flat_1$flag_gastrap))

flat_1$flag_Pi10 <-  ifelse(flat_1$diff_Pi10 >= 0.600, 1, 0)

sum(na.exclude(flat_1$flag_Pi10))

flat_1$flag_AWT <-  ifelse(flat_1$diff_AWT >= 0.189, 1, 0)

sum(na.exclude(flat_1$flag_AWT))


flat_1$flagcount <- flat_1$flag_spiro_ratio + flat_1$flag_fev1 + flat_1$flag_emphys + flat_1$flag_Pi10 + flat_1$flag_gastrap+flat_1$flag_AWT 

table(flat_1$flagcount)

#based on this table, we can see that most people only experience one of these markers if at all but nearly 3-4% of subjects experience two. There are 7 subjects that experience 4/6 markers of increased severity. 
```

In terms of the mortality dataset:


```{r}
length(mort_adjud$vital_status) #number of subjects in the cohort

sum(na.exclude(mort_adjud$vital_status)) #number of deaths in the cohort

summaries(mort_adjud$months_followed_net) #83 months of average follow up - lower 5% was 13

summaries(mort_adjud$days_followed) # mean days followed was 2491.7 (little weird, 2065 was average days between visits - probably artifact from LFU data)


#if we subset to only subjects who died:

mort_dead <- subset(mort_adjud, mort_adjud$vital_status == 1)
summaries(mort_dead$months_followed_net) #52 months of average follow up time, lower 5% was 7

summaries(mort_dead$days_followed) # mean days followed was 1586.3
```


Merging the datasets:


```{r}

fulldata <- merge(flat_1,mort_adjud,by="sid")

fulldata$sid <- as.factor(fulldata$sid)

#subset only to people who had a visit 2 date. 

fulldata$check2 <- ifelse(is.na(fulldata$Visit_Date_P2), 1, 0)

datavisit2 <- subset(fulldata, fulldata$check2 == 0)

#subset only to people who had a marked change in one of the the biomarkers of interest and a visit 2

#datasicker <- subset(fulldata, fulldata$flagcount > 0)


#Pull a random sample of subject ids to evaluate the biomarker progression of the subjects

set.seed(245)
randomsid <- sample(datavisit2$sid, 30, replace=FALSE)
#$randomsidsick <- sample(datasicker$sid, 30, replace=FALSE)

#create binary variable for where the item is true

datavisit2$check <- ifelse(datavisit2$sid %in% randomsid, 1, 0)
#datasicker$check <- ifelse(datasicker$sid %in% randomsidsick, 1, 0)
#subset our dataframe to just have these values:
randomdat <- subset(datavisit2, datavisit2$check == 1)

#randomsick <- subset(datasicker, datasicker$check == 1)
```


Now make all of the necessary plots that show the progression of the different biomarkers. 

Problem: need to change it to long format instead of wide.

 This is for everyone in the data set, even those that don't hit the extra sick markers.
```{r}
widerandom <- randomdat %>% select(sid, Visit_Date_P1, Visit_Date_P2, FEV1_FVC_utah_P1,FEV1_FVC_utah_P2, FEV1_utah_P1,FEV1_utah_P2, FVC_utah_P1,FVC_utah_P2, pctEmph_Thirona_P1, pctEmph_Thirona_P2, AWT_seg_Thirona_P1,  AWT_seg_Thirona_P2,  pctGasTrap_Thirona_P1,pctGasTrap_Thirona_P2, Pi10_Thirona_P1,Pi10_Thirona_P2)

library(reshape)
library(reshape2)
library(ggplot2)
library(magrittr)
library(dplyr)
library(gridExtra)


#great, now it's in long format and I can start creating the graphs that I need. 
longrandom <- reshape(widerandom,idvar='sid', direction='long', 
        varying=list(c(2,3), c(4,5), c(6,7), c(8,9), c(10,11), c(12,13), c(14,15), c(16,17)), #note that these are paired to match eachother
        timevar='visit',
        times=c('p1', 'p2'),
        v.names=c('visitdate','fev1_fvc', 'fev1', 'fvc', 'pct_emph', 'awt', 'pct_gastrap', 'pi10'))


#for practice: we'll just do it on one item first





plotting <-function(x) {
a <- ggplot(x, aes(log(visitdate), fev1_fvc)) + 
  geom_point(color = 'purple') +geom_path(color = 'purple') + scale_x_continuous(breaks=seq(9.75,9.85, 9.96)) + xlab("log(day)") + ylim(0.2,1)
b <- ggplot(x, aes(log(visitdate), fev1)) + 
  geom_point(color = 'red') +geom_path(color = 'red') + scale_x_continuous(breaks=seq(9.75,9.85, 9.96)) + xlab("log(day)")+ ylim(0.5,4.25)
c <- ggplot(x, aes(log(visitdate), fvc)) + 
  geom_point(color='blue') +geom_path(color='blue') + scale_x_continuous(breaks=seq(9.75,9.85, 9.96)) + xlab("log(day)")+ ylim(1.5,5.3)
d <- ggplot(x, aes(log(visitdate), pct_emph)) + 
  geom_point(color='orangered') +geom_path(color='orangered') + scale_x_continuous(breaks=seq(9.75,9.85, 9.96)) + xlab("log(day)") + ylim(0,50)
e <- ggplot(x, aes(log(visitdate), awt)) + 
  geom_point(color = 'forestgreen') +geom_path(color = 'forestgreen') + scale_x_continuous(breaks=seq(9.75,9.85, 9.96)) + xlab("log(day)") + ylim(0.5,1.5)
f <- ggplot(x, aes(log(visitdate), pct_gastrap)) + 
  geom_point(color='magenta') +geom_path(color='magenta') + scale_x_continuous(breaks=seq(9.75,9.85, 9.96)) + xlab("log(day)") + ylim(0,76) 
g <- ggplot(x, aes(log(visitdate), pi10)) + 
  geom_point(color = 'darkblue') +geom_path(color = 'darkblue')  + scale_x_continuous(breaks=seq(9.75,9.85, 9.96)) + xlab("log(day)")+ ylim(1.3,4) 


return(grid.arrange(a,b,c,d,e,f,g, ncol=7))

}





subjects <- unique(longrandom$sid)
for (i in 1:30) {
  x <- as.data.frame(subset(longrandom, longrandom$sid == subjects[i]))
  a <- plotting(x)
  a
}




```



Change variables to fit data structures:

```{r} 
fulldata$gender <- as.factor(fulldata$gender)
fulldata$race <- as.factor(fulldata$race)
fulldata$ccenter_P1 <- as.factor(fulldata$ccenter_P1)
fulldata$ccenter_P2 <- as.factor(fulldata$ccenter_P2)
fulldata$finalGold_P1 <- as.factor(fulldata$finalGold_P1)
fulldata$finalGold_P2 <- as.factor(fulldata$finalGold_P2)
fulldata$EverSmokedCig_P1<- as.factor(fulldata$EverSmokedCig_P1)
fulldata$smoking_status_P1<- as.factor(fulldata$smoking_status_P1)
```

Now for summary tables.

We need to define the population of interest, that is; we want all subjects that either complete both visits or failed to complete both visits because they died before visit 1. 

```{r}
#subjects that died before visit 2:

fulldata$deadearly <- ifelse(is.na(fulldata$Visit_Date_P2)& fulldata$vital_status==1, 1, 0)

#subjects that made it to visit 2:
fulldata$v2attend<- ifelse(!is.na(fulldata$Visit_Date_P2), 1, 0)

#subjects that made it to visit 2 and are known to be dead

fulldata$deadlater <- ifelse(fulldata$v2attend == 1 & fulldata$vital_status ==1, 1, 0)

subjectdata <- subset(fulldata, fulldata$deadearly ==1 | fulldata$v2attend == 1 | fulldata$deadlater == 1)
subjectdata$group1 <- rep(0, 8158)

subjectdata$groupa <- ifelse(subjectdata$deadearly == 1, 1, subjectdata$group1)
subjectdata$groupb <- ifelse(subjectdata$v2attend == 1, 2, subjectdata$groupa)
subjectdata$group <- ifelse(subjectdata$deadlater == 1, 3, subjectdata$groupb)


```




Now, create a general table 1, then a table 1 for each of the other groups.

```{r}
library(tableone)
## Vector of variables to summarize
myVars <- c("Age_P1", "BMI_P1", "ATS_PackYears_P1", "FEV1_FVC_utah_P1", "FEV1_utah_P1", "FVC_utah_P1",
            "pctEmph_Thirona_P1", "AWT_seg_Thirona_P1", "pctGasTrap_Thirona_P1","gender", "race", "EverSmokedCig_P1", "smoking_status_P1",
             "finalGold_P1", "finalGold_P2")
## Vector of categorical variables that need transformation
catVars <- c("gender", "race", "EverSmokedCig_P1", "smoking_status_P1",
             "finalGold_P1", "finalGold_P2")
## Create  TableOne objects
tab2 <- CreateTableOne(vars = myVars, data = subjectdata, factorVars = catVars)
tab2 #values for all patients
summary(tab2)


```

```{r}

subjectdata$group <- as.factor(subjectdata$group)
## Vector of variables to summarize
myVars <- c("Age_P1", "BMI_P1", "ATS_PackYears_P1", "FEV1_FVC_utah_P1", "FEV1_utah_P1", "FVC_utah_P1",
            "pctEmph_Thirona_P1", "AWT_seg_Thirona_P1", "pctGasTrap_Thirona_P1","gender", "race", "EverSmokedCig_P1", "smoking_status_P1",
             "finalGold_P1", "finalGold_P2", "group")
## Vector of categorical variables that need transformation
catVars <- c("gender", "race", "EverSmokedCig_P1", "smoking_status_P1",
             "finalGold_P1", "finalGold_P2", "group")
## Create  TableOne objects
tab3 <- CreateTableOne(vars = myVars, strata = c("group"), data = subjectdata, factorVars = catVars)
tab3 #values for all patients
summary(tab3)


```

```{r}
sum(is.na(flat_1$Visit_Date_P2)) #3962 subjects did not have a visit 2 date
sum(is.na(flat_1$Visit_Date_P1)) #349 subjects do not have a visit 1 date

sum(is.na(flat_1$FEV1_utah_P2)) #4624 patients don't have fev1 for visit 2
sum(is.na(flat_1$FEV1_utah_P1)) #416 patients don't have fev1 for visit 1

sum(is.na(flat_1$FEV1_FVC_utah_P2)) #4625 patients don't have fev1/fvc for visit 2
sum(is.na(flat_1$FEV1_FVC_utah_P1)) #416 patients don't have fev1/fvc for visit 1


sum(is.na(flat_1$pctEmph_Thirona_P2)) #4995 patients don't have emphysema percentage for visit 2
sum(is.na(flat_1$pctEmph_Thirona_P1)) #1072 patients didn't have visit 1 emphysema

sum(is.na(flat_1$AWT_seg_Thirona_P2)) #5000 patients don't have AWT measures for visit 2
sum(is.na(flat_1$AWT_seg_Thirona_P1)) #1073 patients dont' have AWT measures for visit 1

sum(is.na(flat_1$Pi10_Thirona_P2)) #4995 patients don't have Pi10 for visit 2
sum(is.na(flat_1$Pi10_Thirona_P1)) #1072 patients don't have Pi10 values for visit 1


sum(is.na(flat_1$pctGasTrap_Thirona_P2)) #5505 patients don't have gas trapping for visit 2
sum(is.na(flat_1$pctGasTrap_Thirona_P1)) #2435 patients don't have gas trapping for visit 1

```














